---
title: Data Exploration
author: Faysal Shaikh
output: 
    html_document:
        toc: true
        toc_depth: 3
        toc_float: true
---

```{r load_relevant_packages, warning=F}
library(tidyverse)
library(readxl)
library(forecast)
```

This document contains exploratory data analysis on opioid data sources to-be used in this project.

# Opioid Data Sources

```{r specify_data_paths}
data_dir_path <- # uncomment appropriate option below
    # "d:/Codebase/ms-thesis/data/" # laptop
    "f:/Codebase/ms-thesis/data/" # desktop

vdh_ed_visits_fname <- "Drug-Overdose-ED-Visits_Virginia-January-2024.xlsx"
cdc_wonder_overdose_deaths_fname <- # uncomment appopriate option below
    # "outcomes.tsv" # old file version
    "cdc-wonder_final-mcd_1999-2020.tsv" # 1999-2020 final
    # "cdc-wonder_final-mcd_2018-2021.tsv" # 2018-2021 final
    # "cdc-wonder_provisional-mcd_2018-feb2024.tsv" # 2018-feb2024 provisional
```

As of now, we are using 2 types of data sources: 

 - opioid overdose emergency department visits
 - opioid overdose deaths

Below, we perform data exploration on these sources separately.

## Emergency Department Visits

```{r read_vdh_ed_visits_data, message=F, warning=F}
vdh_ed_visits_fullpath <- paste0(data_dir_path, vdh_ed_visits_fname) # precombine fname parts
vdh_ed_visits_sheetslist <- excel_sheets(vdh_ed_visits_fullpath) # get list of worksheet "tabs" for MS Excel workbook file

vdh_ed_visits_nonheroin_opioid <- read_excel(
    path = vdh_ed_visits_fullpath,
    sheet = vdh_ed_visits_sheetslist[5] # locality opioid
)
vdh_ed_visits_heroin <- read_excel(
    path = vdh_ed_visits_fullpath,
    sheet = vdh_ed_visits_sheetslist[6] # locality heroin
)

# preprocessing
dates_vec <- seq(as.Date("2015-01-01"), as.Date("2024-01-01"), "+1 month") %>% format("%B %Y")
vdh_localities_df <- vdh_ed_visits_nonheroin_opioid["OPIOID"][-1,] %>% 
    rename("Locality" = "OPIOID") %>% 
    lapply(function(x) str_replace_all(x, "[^a-zA-Z0-9 ]", "")) %>%  # remove special characters (i.e., double-dagger)
    as.data.frame()

nonheroin_opioid_counts_df_wide <- vdh_localities_df %>% cbind(vdh_ed_visits_nonheroin_opioid[dates_vec][-1,])

nonheroin_opioid_counts_ts_list <- sapply( # ts objects stored in named list
    vdh_localities_df[,], 
    function(x) nonheroin_opioid_counts_df_wide[nonheroin_opioid_counts_df_wide["Locality"]==x][2:110] %>%
        as.numeric() %>% ts(start=2015, end=2024, frequency=12),
    simplify=FALSE, USE.NAMES=TRUE
)

nonheroin_opioid_counts_df <- nonheroin_opioid_counts_df_wide %>% pivot_longer(
    cols = dates_vec,
    names_to = "Month",
    values_to = "Count: Nonheroin Opioid ED Visits"
)

heroin_counts_df_wide <- vdh_localities_df %>% cbind(vdh_ed_visits_heroin[dates_vec][-1,])

heroin_counts_ts_list <- sapply( # ts objects stored in named list
    vdh_localities_df[,], 
    function(x) heroin_counts_df_wide[heroin_counts_df_wide["Locality"]==x][2:110] %>%
        as.numeric() %>% ts(start=2015, end=2024, frequency=12),
    simplify=FALSE, USE.NAMES=TRUE
)

heroin_counts_df <- heroin_counts_df_wide %>% pivot_longer(
    cols = dates_vec,
    names_to = "Month",
    values_to = "Count: Heroin ED Visits"
)
```

According to VDH criteria, [opioid data seems to exclude "heroin"](https://www.vdh.virginia.gov/surveillance-and-investigation/syndromic-surveillance/drug-overdose-case-definition/#:~:text=medication%20or%20drug%2C-,heroin,-%2C%20dope%2C%20speed%20ball) ICD codes; may be relevant to combine opioid and heroin data from VDH sources.

## Overdose Deaths

```{r read_cdc_wonder_overdose_deaths_data, warning=F}
cdc_wonder_overdose_deaths_fullpath <- paste0(data_dir_path, cdc_wonder_overdose_deaths_fname)

cdc_wonder_overdose_deaths_df <- read_tsv(cdc_wonder_overdose_deaths_fullpath, # already in long format
    col_select = c("County", "Year", "Deaths", "Population") # final cause of death version
    # col_select = c("Residence County", "Year", "Deaths", "Population") # provisional cause of death version
) %>% drop_na()

cdc_localities_df <- cdc_wonder_overdose_deaths_df[,1] %>% unique() %>% as.data.frame()

cdc_wonder_overdose_deaths_df_wide <- cdc_wonder_overdose_deaths_df %>% 
    select(c(County, Year, Deaths)) %>% group_by(County) %>% 
    pivot_wider(names_from = Year, values_from = Deaths)

cdc_wonder_overdose_deaths_ts_list <- sapply(
    cdc_localities_df[,], 
    function(x) cdc_wonder_overdose_deaths_df_wide[cdc_wonder_overdose_deaths_df_wide["County"]==x,] %>% 
        as.numeric() %>% ts(start=1999, end=2020), # ensure this is updated to the data
    simplify=FALSE,
    USE.NAMES=TRUE
)
```

I have downloaded 3 different versions of CDC WONDER data: 
 - bridged-race final multiple causes of death (1999-2020), 
 - single-race final multiple cause of death (2018-2021), and
 - provisional cause of death (2018-feb2024).

Data utilized apropriate ICD10 codes for underlying cause of death as drug overdose and multiple causes of death codes involving opioid drugs, as specified by the [KFF](https://www.kff.org/other/state-indicator/opioid-overdose-death-rates/?currentTimeframe=0&sortModel=%7B%22colId%22:%22Location%22,%22sort%22:%22asc%22%7D) and a [SAMHSA resource document](https://mnprc.org/wp-content/uploads/2019/01/using-icd-10-codes-to-assess-opioid-related-overdose-deaths.pdf).

# Data Missingness

## Overdose Emergency Department Visits
```{r tabulate_data_missingness_edvisits}
vdh_ed_visits_heroin
```

## Overdose Deaths
```{r tabulate_data_missingness_deaths}
cdc_wonder_overdose_deaths_df %>% 
    mutate(Deaths = as.numeric(Deaths)) %>% 
    group_by(County) %>% 
    summarize(nonNA_deaths = sum(!is.na(Deaths)), nonzero_nonNA_deaths = sum(!is.na(Deaths) & (Deaths != 0))) %>% 
    arrange(desc(nonzero_nonNA_deaths)) %>% # sort descending by nonzero&nonNA values
    print(n=135) # print all 135 counties
```

# Exploratory Data Visualization

## Overdose Emergency Department Visits

```{r plotting_county_heroin_ed_visits, eval=F}
heroin_counts_df %>% # all data (messy)
    ggplot(aes(x=Month, y=`Count: Heroin ED Visits`, group=Locality)) + 
        geom_line(aes(color=Locality)) + geom_point(aes(color=Locality)) + 
        theme_classic() + theme(legend.position = "none")

heroin_counts_ts_list %>% names() %>% 
    sapply(function(x) ggseasonplot(heroin_counts_ts_list[[x]], polar=TRUE) + theme_classic(), # seasonal polar plot
        simplify=FALSE, USE.NAMES=TRUE
)

heroin_counts_df[heroin_counts_df["Locality"] == "VIRGINIA",] %>% # Virginia state overall
    ggplot(aes(x=Month, y=`Count: Heroin ED Visits`, group=Locality)) + 
        geom_line(aes(color=Locality)) + geom_point(aes(color=Locality)) + 
        theme_classic() + theme(legend.position = "none")

chosen_areas <- c("Alexandria City", "Fairfax County", "Fairfax City")
heroin_counts_df[heroin_counts_df$Locality %in% chosen_areas,] %>% # chosen areas
    ggplot(aes(x=Month, y=`Count: Heroin ED Visits`, group=Locality)) + 
        geom_line(aes(color=Locality)) + geom_point(aes(color=Locality)) + 
        theme_classic() + theme(legend.position = "top")
```

```{r plotting_county_nonheroin_opioid_ed_visits, eval=F}
nonheroin_opioid_counts_df %>% # all data (messy)
    ggplot(aes(x=Month, y=`Count: Nonheroin Opioid ED Visits`, group=Locality)) + 
        geom_line(aes(color=Locality)) + geom_point(aes(color=Locality)) + 
        theme_classic() + theme(legend.position = "none")

nonheroin_opioid_counts_ts_list %>% names() %>% 
    sapply(function(x) ggseasonplot(nonheroin_opioid_counts_ts_list[[x]], polar=TRUE) + theme_classic(), # seasonal polar plot
        simplify=FALSE, USE.NAMES=TRUE
)

nonheroin_opioid_counts_df[nonheroin_opioid_counts_df["Locality"] == "VIRGINIA",] %>% # Virginia state overall
    ggplot(aes(x=Month, y=`Count: Nonheroin Opioid ED Visits`, group=Locality)) + 
        geom_line(aes(color=Locality)) + geom_point(aes(color=Locality)) + 
        theme_classic() + theme(legend.position = "none")

chosen_areas <- c("Alexandria City", "Fairfax County", "Fairfax City")
nonheroin_opioid_counts_df[nonheroin_opioid_counts_df$Locality %in% chosen_areas,] %>% # chosen areas
    ggplot(aes(x=Month, y=`Count: Nonheroin Opioid ED Visits`, group=Locality)) + 
        geom_line(aes(color=Locality)) + geom_point(aes(color=Locality)) + 
        theme_classic() + theme(legend.position = "top")
```

## Overdose Deaths

```{r plotting_county_opioid_overdose_deaths, eval=F}
cdc_wonder_overdose_deaths_df %>% # all data (messy)
    ggplot(aes(x=Year, y=Deaths, group=County)) + 
        geom_line(aes(color=County)) + geom_point(aes(color=County)) + 
        theme_classic() + theme(legend.position = "none")

# yearly data are not seasonal (no seasonal polar plot)

chosen_areas <- c("Alexandria city, VA", "Fairfax County, VA", "Fairfax city, VA") # "County" uppercase and "city" lowercase
cdc_wonder_overdose_deaths_df[cdc_wonder_overdose_deaths_df$County %in% chosen_areas,] %>% # chosen areas
    ggplot(aes(x=Year, y=Deaths, group=County)) + 
        geom_line(aes(color=County)) + geom_point(aes(color=County)) + 
        theme_classic() + theme(legend.position = "top")
```


# Remaining Action Items
 - solve `readRDS(pfile): error reading the file` warnings (50+) issue in [first chunk](#load_relevant_packages) (without simply `warning=F` flagging the chunk)
 - properly describe & reference CDC WONDER data selection: underlying causes as overdose and multiple causes to include all classes of opioid drugs
 - describe missingness/suppression/etc of VA counties for each data source
 - merge CDC WONDER data files: evaluate merge based on # of mismatches b/w overlapping years (2018-2020)
    - merge 1: 1999-2020 final (bridged-race) + 2018-2021 final (single-race)
    - merge 2: merge 1 + 2018-feb2024 provisional
    - keep merge 1 and merge 2 separate, for final versus provisional analyses
 - determine & document inclusion/exclusion criteria for county-trajectories
 - visualize all-county (meeting inclusion criteria) overdose emergency department visits
 - visualize all-county (meeting inclusion criteria) overdose deaths
 - on a per-county basis, visualize all counties' (meeting inclusion criteria) overdose emergency department visits and overdose deaths on the same plot (vertical axis: "cases" rather than unique individuals, etc.)
 - find & visualize relevant national data on this topic