---
title: Data Exploration
author: Faysal Shaikh
output: 
    html_document:
        toc: true
        toc_depth: 3
        toc_float: true
---

```{r load relevant packages}
library(tidyverse)
library(readxl)
library(forecast)
```

This document contains exploratory data analysis on opioid data sources to-be used in this project.

# Opioid Data Sources

```{r specify data paths}
data_dir_path <- "d:/Codebase/ms-thesis/data/"

vdh_ed_visits_fname <- "Drug-Overdose-ED-Visits_Virginia-January-2024.xlsx"
cdc_wonder_overdose_deaths_fname <- "outcomes.tsv"
```

As of now, we are using 2 types of data sources: 

 - opioid overdose emergency department visits
 - opioid overdose deaths

Below, we perform data exploration on these sources separately.

## Emergency Department Visits

```{r read vdh ed visits data, message=F, warning=F}
vdh_ed_visits_fullpath <- paste0(data_dir_path, vdh_ed_visits_fname) # precombine fname parts
vdh_ed_visits_sheetslist <- excel_sheets(vdh_ed_visits_fullpath) # get list of worksheet "tabs" for MS Excel workbook file

vdh_ed_visits_nonheroin_opioid <- read_excel(
    path = vdh_ed_visits_fullpath,
    sheet = vdh_ed_visits_sheetslist[5] # locality opioid
)
vdh_ed_visits_heroin <- read_excel(
    path = vdh_ed_visits_fullpath,
    sheet = vdh_ed_visits_sheetslist[6] # locality heroin
)

# preprocessing
dates_vec <- seq(as.Date("2015-01-01"), as.Date("2024-01-01"), "+1 month") %>% format("%B %Y")
vdh_localities_df <- vdh_ed_visits_nonheroin_opioid["OPIOID"][-1,] %>% 
    rename("Locality" = "OPIOID") %>% 
    lapply(function(x) str_replace_all(x, "[^a-zA-Z0-9 ]", "")) %>%  # remove special characters (i.e., double-dagger)
    as.data.frame()

nonheroin_opioid_counts_df_wide <- vdh_localities_df %>% cbind(vdh_ed_visits_nonheroin_opioid[dates_vec][-1,])

nonheroin_opioid_counts_ts_list <- sapply( # ts objects stored in named list
    vdh_localities_df[,], 
    function(x) nonheroin_opioid_counts_df_wide[nonheroin_opioid_counts_df_wide["Locality"]==x][2:110] %>%
        as.numeric() %>% ts(start=2015, end=2024, frequency=12),
    simplify=FALSE, USE.NAMES=TRUE
)

nonheroin_opioid_counts_df <- nonheroin_opioid_counts_df_wide %>% pivot_longer(
    cols = dates_vec,
    names_to = "Month",
    values_to = "Count: Nonheroin Opioid ED Visits"
)

heroin_counts_df_wide <- vdh_localities_df %>% cbind(vdh_ed_visits_heroin[dates_vec][-1,])

heroin_counts_ts_list <- sapply( # ts objects stored in named list
    vdh_localities_df[,], 
    function(x) heroin_counts_df_wide[heroin_counts_df_wide["Locality"]==x][2:110] %>%
        as.numeric() %>% ts(start=2015, end=2024, frequency=12),
    simplify=FALSE, USE.NAMES=TRUE
)

heroin_counts_df <- heroin_counts_df_wide %>% pivot_longer(
    cols = dates_vec,
    names_to = "Month",
    values_to = "Count: Heroin ED Visits"
)
```

According to VDH criteria, [opioid data seems to exclude "heroin"](https://www.vdh.virginia.gov/surveillance-and-investigation/syndromic-surveillance/drug-overdose-case-definition/#:~:text=medication%20or%20drug%2C-,heroin,-%2C%20dope%2C%20speed%20ball) ICD codes; may be relevant to combine opioid and heroin data from VDH sources.

```{r plotting county heroin ed visits}
heroin_counts_df %>% # all data (messy)
    ggplot(aes(x=Month, y=`Count: Heroin ED Visits`, group=Locality)) + 
        geom_line(aes(color=Locality)) + geom_point(aes(color=Locality)) + 
        theme_classic() + theme(legend.position = "none")

heroin_counts_ts_list %>% names() %>% 
    sapply(function(x) ggseasonplot(heroin_counts_ts_list[[x]], polar=TRUE) + theme_classic(), # seasonal polar plot
        simplify=FALSE, USE.NAMES=TRUE
)

heroin_counts_df[heroin_counts_df["Locality"] == "VIRGINIA",] %>% # Virginia state overall
    ggplot(aes(x=Month, y=`Count: Heroin ED Visits`, group=Locality)) + 
        geom_line(aes(color=Locality)) + geom_point(aes(color=Locality)) + 
        theme_classic() + theme(legend.position = "none")

chosen_areas <- c("Alexandria City", "Fairfax County", "Fairfax City")
heroin_counts_df[heroin_counts_df$Locality %in% chosen_areas,] %>% # chosen areas
    ggplot(aes(x=Month, y=`Count: Heroin ED Visits`, group=Locality)) + 
        geom_line(aes(color=Locality)) + geom_point(aes(color=Locality)) + 
        theme_classic() + theme(legend.position = "top")
```

```{r plotting county nonheroin opioid ed visits}
nonheroin_opioid_counts_df %>% # all data (messy)
    ggplot(aes(x=Month, y=`Count: Nonheroin Opioid ED Visits`, group=Locality)) + 
        geom_line(aes(color=Locality)) + geom_point(aes(color=Locality)) + 
        theme_classic() + theme(legend.position = "none")

nonheroin_opioid_counts_ts_list %>% names() %>% 
    sapply(function(x) ggseasonplot(nonheroin_opioid_counts_ts_list[[x]], polar=TRUE) + theme_classic(), # seasonal polar plot
        simplify=FALSE, USE.NAMES=TRUE
)

nonheroin_opioid_counts_df[nonheroin_opioid_counts_df["Locality"] == "VIRGINIA",] %>% # Virginia state overall
    ggplot(aes(x=Month, y=`Count: Nonheroin Opioid ED Visits`, group=Locality)) + 
        geom_line(aes(color=Locality)) + geom_point(aes(color=Locality)) + 
        theme_classic() + theme(legend.position = "none")

chosen_areas <- c("Alexandria City", "Fairfax County", "Fairfax City")
nonheroin_opioid_counts_df[nonheroin_opioid_counts_df$Locality %in% chosen_areas,] %>% # chosen areas
    ggplot(aes(x=Month, y=`Count: Nonheroin Opioid ED Visits`, group=Locality)) + 
        geom_line(aes(color=Locality)) + geom_point(aes(color=Locality)) + 
        theme_classic() + theme(legend.position = "top")
```

## Overdose Deaths

```{r read cdc wonder overdose deaths data, warning=F}
cdc_wonder_overdose_deaths_fullpath <- paste0(data_dir_path, cdc_wonder_overdose_deaths_fname)

cdc_wonder_overdose_deaths_df <- read_tsv(cdc_wonder_overdose_deaths_fullpath, # already in long format
    col_select = c("County", "Year", "Deaths", "Population")
) %>% drop_na()

cdc_localities_df <- cdc_wonder_overdose_deaths_df[,1] %>% unique() %>% as.data.frame()

cdc_wonder_overdose_deaths_df_wide <- cdc_wonder_overdose_deaths_df %>% 
    select(c(County, Year, Deaths)) %>% group_by(County) %>% 
    pivot_wider(names_from = Year, values_from = Deaths)

cdc_wonder_overdose_deaths_ts_list <- sapply(
    cdc_localities_df[,], 
    function(x) cdc_wonder_overdose_deaths_df_wide[cdc_wonder_overdose_deaths_df_wide["County"]==x,] %>% 
        as.numeric() %>% ts(start=2010, end=2019),
    simplify=FALSE,
    USE.NAMES=TRUE
)
```

As of Wed. 02/28, having issues accessing CDC WONDER data.  In the meantime, have decided to use old CDC WONDER data we had access to (`outcomes.tsv`).

```{r plotting county opioid overdose deaths}
cdc_wonder_overdose_deaths_df %>% # all data (messy)
    ggplot(aes(x=Year, y=Deaths, group=County)) + 
        geom_line(aes(color=County)) + geom_point(aes(color=County)) + 
        theme_classic() + theme(legend.position = "none")

# yearly data are not seasonal (no seasonal polar plot)

chosen_areas <- c("Alexandria city, VA", "Fairfax County, VA", "Fairfax city, VA") # "County" uppercase and "city" lowercase
cdc_wonder_overdose_deaths_df[cdc_wonder_overdose_deaths_df$County %in% chosen_areas,] %>% # chosen areas
    ggplot(aes(x=Year, y=Deaths, group=County)) + 
        geom_line(aes(color=County)) + geom_point(aes(color=County)) + 
        theme_classic() + theme(legend.position = "top")
```

# Data Missingness